---
title: "Binary AddiVortes: probit classification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary AddiVortes: probit classification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Binary AddiVortes is the classification variant of Stone and Gosling
(2025): P(y = 1 | x) = Phi(c + f(x)), fitted with the latent-variable
augmentation of Albert and Chib (1993), as BART's `pbart` fits its
probit model. It is selected by the `probit_outcome()` family. The offset c
defaults to Phi^-1(ybar), resolved at fit.

```{r}
library(thiessen)

set.seed(1)
n <- 300
x <- cbind(runif(n), runif(n))
label <- factor(ifelse(x[, 1] + 0.3 * x[, 2] + rnorm(n, sd = 0.2) > 0.6,
                       "yes", "no"))

control <- thiessen_control(
  outcome = probit_outcome(),
  tessellations = 25,
  general_params = general_params(burn_in = 100, draws = 200)
)
fit <- thiessen(x, label, control, seed = 1)
fit
```

A two-level factor response becomes 0 and 1 with the first level as the
zero, as `glm` treats one; the levels are kept on the fit. `predict()`
returns the posterior mean probability of the second level, and
`predict(type = "latent")` the latent mean c + f(x) per draw:

```{r}
probabilities <- predict(fit, x)
range(probabilities)
mean((probabilities > 0.5) == (label == "yes"))
```

Credible intervals are on the probability scale:

```{r}
head(predict(fit, x, interval = "credible", level = 0.9))
```

## What the model fixes

The latent scale is 1 for identification, so `sigma()` is empty, a
variance ensemble is not available, and `predict(interval =
"prediction")` errors: a two-point distribution has no continuous
predictive interval. The likelihood in `log_lik()` is Bernoulli, and
`fit$in_sample_rmse` is the root Brier score.

## References

Albert, J. H. and Chib, S. (1993). Bayesian analysis of binary and
polychotomous response data. *Journal of the American Statistical
Association* 88(422), 669-679.

Stone, A. and Gosling, J. P. (2025). AddiVortes: (Bayesian) additive
Voronoi tessellations. *Journal of Computational and Graphical
Statistics* 34(3), 859-871.
