---
title: "The control surface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The control surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The configuration of a fit has four parts, and `thiessen_control()`
holds one of each. Every name here is a name in the core's stored
configuration and in the Python package, so one description serves all
three surfaces. The outcome constructors are the exception: they carry
an `_outcome` suffix in R, and the name the configuration stores is
unchanged, so a fit saved by one surface is read by another. The full
mapping is [`docs/parity.md`](https://github.com/l-thomson/thiessen/blob/dev/docs/parity.md).

```{r}
library(thiessen)

thiessen_control()
```

## The outcome family

`outcome` selects the observation model:
`gaussian_outcome(nu, q)` for continuous responses, carrying the sigma^2
prior's degrees of freedom and calibration quantile, and
`probit_outcome(offset)` for binary ones. The suffix keeps
`gaussian_outcome()` from masking `stats::gaussian()`, which `glm()` takes,
and both families carry it rather than only the one that would clash.
Attaching `variance_params` with a positive tessellation count extends the
Gaussian family to the heteroscedastic model. `outcome` left `NULL`, the
default, takes the family from the response at fit: a numeric vector the
Gaussian family and a two-level factor the probit family, with the
experimental families selected by their own response shapes (the
[experimental](experimental.html) page lists them). A family
named here is checked against the response and a mismatch is an error
naming both.

```{r}
gaussian_outcome(nu = 3)
probit_outcome()
```

## The ensembles

`mean_params` describes the ensemble behind the average and
`variance_params`, when given, the ensemble behind the spread. Each is a
`term_params()` group: the tessellation count, the cell-value prior
spread `k`, the cell-count prior rate `lambda_c`, and two nested groups.

```{r}
term_params(tessellations = 200, k = 3, lambda_c = 25)
```

`geometry_params()` is the covariate space: the `metric` of each column
and the centre-coordinate scale `sigma_c`. A default build reaches three
metrics, `"euclidean"`, `"categorical"` and a labelled sphere; the rest
the core carries are experimental and are rejected without the opt-in
build. A Euclidean column is min-max scaled to [-0.5, 0.5] over its
training range inside the sampler, which is what makes one distance
comparable across columns; a categorical or spherical column is not
scaled, and `sigma_c` is on the scaled coordinate, so 1 is the full
range of a Euclidean column. `structure_params()` is the
covariate-inclusion prior: `omega / p` is the prior probability of
including a covariate, resolved to `min(3, p)` at fit when unset. The
two ensembles share one covariate space: set these on `mean_params` and
they apply to `variance_params` as well.

```{r}
term_params(
  geometry = geometry_params(metric = list("euclidean", "categorical")),
  structure = structure_params(omega = 2)
)
```

## The sweep schedule

`general_params()` carries the run lengths: `burn_in`, `draws`,
`thinning`, and `prior_only`, which switches the likelihood off so the
chain draws from the prior. The chains and threads of a fit are
arguments of `thiessen()` rather than of the schedule, since they change
nothing about the model.

```{r}
general_params(burn_in = 500, draws = 2000, thinning = 2)
```

## The shortcut, and validation

`thiessen_control(tessellations = 200)` sets the mean ensemble's size
without spelling the group; it is the single number most fits tune, and
it is the only flat argument. Every value is validated by the core at
construction, with the reason kept:

```{r, error = TRUE}
thiessen_control(outcome = gaussian_outcome(q = 1.5))
thiessen_control(
  outcome = probit_outcome(),
  variance_params = term_params(tessellations = 40)
)
```

An unknown name in any group is an ordinary unused-argument error, since
each group is a function with a fixed signature:

```{r, error = TRUE}
term_params(zeta = 1)
```

## Defaults

An argument left at its default gives the core's default, so
`thiessen_control()` is the published configuration. The core resolves the
configuration at fit and the fit carries what it resolved, so the values
below are read from the core rather than repeated here.

```{r}
n <- 40
x <- cbind(
  a = seq(0, 1, length.out = n),
  b = rep(c(0, 0.25, 0.5, 0.75), length.out = n),
  c = rep(c(0, 1), length.out = n)
)
y <- 2 * (x[, 1] - 0.5)^2 + 0.5 * x[, 2]

resolved <- thiessen(x, y, thiessen_control(
  general_params = general_params(burn_in = 1, draws = 1)
), seed = 1, chains = 1)$control
mean_params <- resolved$mean_params

c(
  tessellations = mean_params$tessellations,
  k = mean_params$k,
  lambda_c = mean_params$lambda_c,
  sigma_c = mean_params$geometry$sigma_c,
  omega = mean_params$structure$omega,
  nu = resolved$outcome$nu,
  q = resolved$outcome$q
)
```

`omega` is the one that depends on the data, resolving to `min(3, p)`,
three of the three covariates above. The one default that departs from the
paper is `lambda_c`: Stone and Gosling (2025), s. 2.3, report 25, and CRAN
AddiVortes takes 5 from 0.6.8 onward; this package follows the
implementation, and `term_params(lambda_c = 25)` is the paper's setting.
The [priors](priors.html) page draws from the prior at three settings
of each.
