---
title: "The control surface"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The control surface}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The configuration of a fit has four parts, and `thiessen_control()`
holds one of each. Every name is a name in the core's stored
configuration and in the Python package, so one description serves all
three surfaces; the full mapping is `docs/parity.md` in the repository.

```{r}
library(thiessen)

thiessen_control()
```

## The outcome family

`outcome` selects the observation model, the `glm(family = )` idiom:
`gaussian(nu, q)` for continuous responses, carrying the sigma^2 prior's
degrees of freedom and calibration quantile, and `probit(offset)` for
binary ones. Attaching `variance_params` with a positive tessellation
count extends the Gaussian family to the heteroscedastic model.

```{r}
gaussian(nu = 3)
probit()
```

## The ensembles

`mean_params` describes the ensemble behind the average and
`variance_params`, when given, the ensemble behind the spread. Each is a
`term_params()` group: the tessellation count, the cell-value prior
spread `k`, the cell-count prior rate `lambda_c`, and two nested groups.

```{r}
term_params(tessellations = 200, k = 3, lambda_c = 25)
```

`geometry_params()` is the covariate space: the `metric` of each column
(`"euclidean"`, `"categorical"`, or a labelled sphere) and the
centre-coordinate scale `sigma_c`. `structure_params()` is the
covariate-inclusion prior: `omega / p` is the prior probability of
including a covariate, resolved to `min(3, p)` at fit when unset. The
two ensembles share one covariate space: set these on `mean_params` and
they apply to `variance_params` as well.

```{r}
term_params(
  geometry = geometry_params(metric = list("euclidean", "categorical")),
  structure = structure_params(omega = 2)
)
```

## The sweep schedule

`general_params()` carries the run lengths: `burn_in`, `draws`,
`thinning`, and `prior_only`, which switches the likelihood off so the
chain draws from the prior.

```{r}
general_params(burn_in = 500, draws = 2000, thinning = 2)
```

## The shortcut, and validation

`thiessen_control(tessellations = 200)` sets the mean ensemble's size
without spelling the group; it is the single number most fits tune, and
it is the only flat argument. Every value is validated by the core at
construction, with the reason kept:

```{r, error = TRUE}
thiessen_control(outcome = gaussian(q = 1.5))
thiessen_control(
  outcome = probit(),
  variance_params = term_params(tessellations = 40)
)
```

An unknown name in any group is an ordinary unused-argument error, since
each group is a function with a fixed signature:

```{r, error = TRUE}
term_params(zeta = 1)
```

## Defaults

An argument left at its default gives the core's default, so
`thiessen_control()` is the published configuration: 200 tessellations,
`k = 3`, `lambda_c = 5`, `nu = 6`, `q = 0.85`, `sigma_c = 0.8`,
`omega = min(3, p)`. The one default that departs from the paper is
`lambda_c`: Stone and Gosling (2025), s. 2.3, report 25, and CRAN
AddiVortes takes 5 from 0.6.8 onward; this package follows the
implementation, and `term_params(lambda_c = 25)` is the paper's
setting.
