---
title: "H-AddiVortes: heteroscedastic regression"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{H-AddiVortes: heteroscedastic regression}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

H-AddiVortes is the heteroscedastic variant of Stone and Gosling
(2025): y = f(x) + e with e ~ N(0, s^2(x)), where the variance s^2(x) is
a multiplicative ensemble of inverse-gamma variance tessellations, the
structure of HBART (Pratola, Chipman, George and McCulloch, 2020). It is
selected by attaching a variance ensemble to the Gaussian family: a
positive tessellation count on `variance_params`. The paper's count is
40.

```{r}
library(thiessen)

set.seed(1)
n <- 300
x <- cbind(runif(n), runif(n))
# The noise scale grows with the first covariate.
y <- x[, 1] + rnorm(n, sd = 0.05 + 0.3 * x[, 1])

control <- thiessen_control(
  mean_params = term_params(tessellations = 25),
  variance_params = term_params(tessellations = 10),
  general_params = general_params(burn_in = 100, draws = 200)
)
fit <- thiessen(x, y, control, seed = 1)
fit
```

`predict(type = "variance")` returns s^2(x) per draw; its posterior mean
recovers the rising noise scale:

```{r}
variance <- colMeans(predict(fit, x, type = "variance"))
order <- order(x[, 1])
c(
  low_x = mean(variance[head(order, 50)]),
  high_x = mean(variance[tail(order, 50)])
)
```

Predictive intervals widen where the variance is high, which is the
point of the model:

```{r}
interval <- predict(fit, x, interval = "prediction", level = 0.9)
width <- interval[, 2] - interval[, 1]
c(
  low_x = mean(width[head(order, 50)]),
  high_x = mean(width[tail(order, 50)])
)
```

## What is shared and what is separate

The two ensembles declare one covariate space: `geometry` and
`structure` set on `mean_params` apply to `variance_params` as well.
The variance ensemble needs a sampled sigma^2 to carry, so it requires
the Gaussian family with nu > 2, and it is not available under the
probit family, whose latent scale is fixed at 1 for identification.
`sigma()` is empty under this model; the spread is `predict(type =
"variance")`.

## References

Pratola, M. T., Chipman, H. A., George, E. I. and McCulloch, R. E.
(2020). Heteroscedastic BART via multiplicative regression trees.
*Journal of Computational and Graphical Statistics* 29(2), 405-417.

Stone, A. and Gosling, J. P. (2025). AddiVortes: (Bayesian) additive
Voronoi tessellations. *Journal of Computational and Graphical
Statistics* 34(3), 859-871.
