---
title: "Working with the posterior"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with the posterior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
suggested <- c("posterior", "loo", "bayesplot", "tidybayes", "ggplot2")
available <- all(vapply(
  suggested, requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = available,
  fig.width = 7, fig.height = 4, dpi = 96, out.width = "100%"
)
if (available) ggplot2::theme_set(ggplot2::theme_minimal())
```

A fit carries its draws, so the established Bayesian tooling takes one
without adaptation: posterior for the draws and their summaries,
bayesplot for the figures, loo for cross-validation and tidybayes for
tidy data frames of draws. Nothing on this page is a method of this
package beyond the generics it registers. Every package used here sits
in `Suggests`, so each chunk is skipped where one is absent; a rendering
with no output below is that, not a failure.

```{r fit}
library(thiessen)
library(posterior)
library(ggplot2)

set.seed(1)
n <- 150
x <- cbind(a = runif(n), b = runif(n))
y <- 2 * (x[, "a"] - 0.5)^2 + 0.5 * x[, "b"] + rnorm(n, sd = 0.1)

fit <- thiessen(x, y, seed = 1)
```

## The draws

`as_draws_df()` and `as_draws_array()` return the draws in posterior's
own formats, chains kept separate.

```{r draws}
draws <- as_draws_df(fit)
c(draws = ndraws(draws), chains = nchains(draws))
dim(as_draws_array(fit))
```

The variables are the mean function at each training row, `mu[i]`, the
noise scale `sigma`, and two structural counts: `cell_count`, the mean
cells per tessellation, and `dimension_count`, the mean covariates in
use per tessellation.

```{r variables}
grep("^mu\\[", variables(draws), invert = TRUE, value = TRUE)
```

`summarise_draws()` gives the usual table, R-hat and both effective
sample sizes included.

```{r summarise}
summarise_draws(subset_draws(draws, variable = c("sigma", "cell_count")))
```

## Three predictive quantities

The three are distinct and easy to confuse.

- `posterior_epred()` is the mean function, one row per draw: E[y | x]
  without observation noise.
- `posterior_predict()` adds the noise, so it draws replicate
  observations. Under the probit family it returns labels.
- `predictive_interval()` summarises `posterior_predict()` into central
  intervals, one row per observation.

```{r predictive}
dim(posterior_epred(fit, x))
dim(posterior_predict(fit, x))
head(predictive_interval(fit, newdata = x, prob = 0.9), 3)
```

`predict()` reaches the same quantities through one argument, and
`predict(type = "draws")` is `posterior_epred()`.

## bayesplot

bayesplot takes the draws object as it stands. Trace plots of the
scalar variables, one line per chain:

```{r trace}
bayesplot::mcmc_trace(as_draws_array(fit), pars = c("sigma", "cell_count"))
```

A posterior predictive check compares the observed response with
replicates from `posterior_predict()`: the density of `y` over the
densities of fifty replicate data sets.

```{r ppc}
replicates <- posterior_predict(fit)
bayesplot::ppc_dens_overlay(y, replicates[1:50, ])
```

`mcmc_areas()`, `mcmc_dens()` and `mcmc_intervals()` take the same
draws object for distributional displays.

## Cross-validation with loo

`log_lik()` returns the pointwise log-likelihood, draws by
observations, which is what `loo::loo()` expects.

```{r loo}
log_likelihood <- log_lik(fit)
dim(log_likelihood)

estimate <- loo::loo(log_likelihood)
estimate$estimates
```

Read the Pareto k diagnostic before the estimate. The importance
sampling behind PSIS-LOO fails for an observation whose k exceeds 0.7:

```{r pareto}
k <- estimate$diagnostics$pareto_k
c(observations = length(k), unreliable = sum(k > 0.7))
```

Where that count is more than a handful, the estimate is not
trustworthy and the answer is more draws, not a different estimator.
Two fits are compared with `loo::loo_compare()` on their loo objects.

## tidybayes

`tidy_draws()` works on the fit directly, and `spread_draws()` on the
frame it returns, so the draws of `mu[i]` for chosen rows come out as a
long data frame ready for ggplot2.

```{r tidybayes}
tidy <- tidybayes::tidy_draws(fit)
dim(tidy)

rows <- order(x[, "a"])[seq(1, n, length.out = 8)]
mu <- tidybayes::spread_draws(tidy, mu[i])
mu <- mu[mu$i %in% rows, ]
head(mu, 3)
```

The posterior of the mean function at those rows, as point and
interval summaries, against the noise-free truth:

```{r pointinterval}
mu$a <- x[mu$i, "a"]
truth <- data.frame(a = x[rows, "a"],
                    f = 2 * (x[rows, "a"] - 0.5)^2 + 0.5 * x[rows, "b"])
ggplot(mu, aes(x = a, y = mu)) +
  tidybayes::stat_pointinterval(colour = "steelblue") +
  geom_point(aes(y = f), data = truth, shape = 4, size = 3) +
  labs(y = "mu[i]")
```

## References

Gabry, J., Simpson, D., Vehtari, A., Betancourt, M. and Gelman, A.
(2019). Visualization in Bayesian workflow. *Journal of the Royal
Statistical Society Series A* 182(2), 389-402.
doi:10.1111/rssa.12378

Vehtari, A., Gelman, A. and Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics
and Computing* 27(5), 1413-1432. doi:10.1007/s11222-016-9696-1

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B. and Buerkner, P.-C.
(2021). Rank-normalization, folding, and localization: an improved R-hat
for assessing convergence of MCMC. *Bayesian Analysis* 16(2), 667-718.
doi:10.1214/20-BA1221
