---
title: "Working with the posterior"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with the posterior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
suggested <- c("posterior", "loo", "bayesplot", "tidybayes")
available <- all(vapply(
  suggested, requireNamespace, logical(1), quietly = TRUE
))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", eval = available,
  fig.width = 6, fig.height = 3, dpi = 96
)
```

Every package this article uses sits in `Suggests`, so each chunk is
skipped where one of posterior, loo, bayesplot or tidybayes is absent. A
rendering with no output below is that, not a failure.

A fit carries its draws, so the established Bayesian tooling takes one
without adaptation. Nothing here is a method of this package beyond the
generics it registers.

```{r fit}
library(thiessen)
library(posterior)

set.seed(1)
n <- 120
x <- cbind(runif(n), runif(n))
y <- 2 * (x[, 1] - 0.5)^2 + 0.5 * x[, 2] + rnorm(n, sd = 0.1)
control <- thiessen_control(
  tessellations = 20,
  general_params = general_params(burn_in = 100, draws = 200)
)

fit <- thiessen(x, y, control, chains = 2, seed = 1)
```

The short schedule here never meets the convergence thresholds, so the
fit warns. That is the warning working, not a problem with the code
below.

## The draws

`as_draws_df()` and `as_draws_array()` return the draws in posterior's
own formats, chains kept separate.

```{r draws}
draws <- as_draws_df(fit)
c(draws = ndraws(draws), chains = nchains(draws))
dim(as_draws_array(fit))
```

The variables are the mean function at each training row, `mu[i]`, the
noise scale `sigma`, and two structural counts: `cell_count`, the cells
across the ensemble, and `dimension_count`, the covariates in use.

```{r variables}
grep("^mu\\[", variables(draws), invert = TRUE, value = TRUE)
```

`summarise_draws()` gives the usual table, R-hat and both effective
sample sizes included.

```{r summarise}
summarise_draws(subset_draws(draws, variable = c("sigma", "cell_count")))
```

## Three predictive quantities

The three are distinct and easy to confuse.

- `posterior_epred()` is the mean function, one row per draw: E[y | x]
  without observation noise.
- `posterior_predict()` adds the noise, so it draws replicate
  observations. Under the probit family it returns labels.
- `predictive_interval()` summarises `posterior_predict()` into central
  intervals, one row per observation.

```{r predictive}
dim(posterior_epred(fit, x))
dim(posterior_predict(fit, x))
head(predictive_interval(fit, newdata = x, prob = 0.9), 3)
```

`predict()` reaches the same quantities through one argument, and
`predict(type = "draws")` is `posterior_epred()`.

## Trace plots

bayesplot takes the draws object as it stands.

```{r trace}
bayesplot::mcmc_trace(as_draws_array(fit), pars = c("sigma", "cell_count"))
```

`thiessen_diagnostics()` returns the same content as a per-draw data
frame for a plot of your own. It is not for printing: a bare `print()`
dumps every draw.

## Cross-validation with loo

`log_lik()` returns the pointwise log-likelihood, draws by
observations, which is what `loo::loo()` expects.

```{r loo}
log_likelihood <- log_lik(fit)
dim(log_likelihood)

estimate <- loo::loo(log_likelihood)
estimate$estimates
```

Read the Pareto k diagnostic before the estimate. The importance
sampling behind PSIS-LOO fails for an observation whose k exceeds 0.7,
and a short schedule on a small data set produces several:

```{r pareto}
k <- estimate$diagnostics$pareto_k
c(observations = length(k), unreliable = sum(k > 0.7))
```

Where that count is more than a handful, the estimate is not
trustworthy and the answer is more draws, not a different estimator.

## tidybayes

`tidy_draws()` works on the fit directly, and `spread_draws()` on the
frame it returns.

```{r tidybayes}
tidy <- tidybayes::tidy_draws(fit)
dim(tidy)

head(tidybayes::spread_draws(tidy, sigma), 3)
```

## References

Vehtari, A., Gelman, A. and Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics
and Computing* 27(5), 1413-1432. doi:10.1007/s11222-016-9696-1

Vehtari, A., Gelman, A., Simpson, D., Carpenter, B. and Buerkner, P.-C.
(2021). Rank-normalization, folding, and localization: an improved R-hat
for assessing convergence of MCMC. *Bayesian Analysis* 16(2), 667-718.
doi:10.1214/20-BA1221
