---
title: "The sampler API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The sampler API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.width = 7, fig.height = 4, dpi = 96, out.width = "100%"
)
ggplot2::theme_set(ggplot2::theme_minimal())
```

An outcome family, a censoring scheme or an imputation scheme this
package does not ship is written in R against `thiessen_sampler()`, with
no Rust and no recompilation. It hands over the Gibbs loop one sweep at
a time and allows the response to be rewritten between sweeps, which is
what a latent-Gaussian data augmentation needs. This page drives a fit
by hand, builds a censored-response model in a dozen lines, and
reimplements the probit family to show that the loop reaches a shipped
model.

The seven verbs are stable. The experimental badge on the object covers
additions to it and changes to what a verb returns beyond its documented
contract.

It follows the updatable sampler object of dbarts and the low-level
interface of stochtree: construct with the configuration, the data and a
seed, then drive the loop yourself. Burn-in and thinning are the
caller's loop.

## The contracts

The response is on the caller's scale through an affine map frozen at
construction, so a response outside the training range is legitimate;
that is what makes censoring and imputation work. The sampler owns its
RNG, seeded at construction with the chain that `thiessen()` would run
first, so driving the configured schedule by hand reproduces a one-chain
fit bit for bit. The loop cannot rewire tessellation membership or cell
internals.

## What is reachable, and what is not

`$set_response()` is the only setter on the object, so a model is
buildable here exactly when it is a Gaussian regression on a response
you can rewrite each sweep.

Reachable, those being the latent-Gaussian data augmentations: probit,
tobit, accelerated failure time, interval-censored and ordinal. The
first is worked below, and the second is the censoring example below
under another name. Four of the five are outcome families the core
carries behind its experimental feature, so this loop is the answer when
a release does not expose the family you need.

Not reachable: an augmentation needing per-observation weights, so
logistic through Polya-Gamma, a Student-t scale mixture, and negative
binomial, because `$noise_variances()` reads the noise and nothing sets
it. Nor the geometry, tessellation membership, cell internals, the
covariate-inclusion prior or the proposals.

Parameters sampled in your own loop, cutpoints for instance, are not in
`$finish()`'s draws. Keep and diagnose those yourself.

## A fit as its own loop

```{r}
library(thiessen)
library(ggplot2)

set.seed(1)
n <- 120
x <- cbind(a = runif(n), b = runif(n))
y <- x[, "a"] + rnorm(n, sd = 0.1)
control <- thiessen_control(
  tessellations = 10,
  general_params = general_params(burn_in = 50, draws = 100)
)

sampler <- thiessen_sampler(x, y, control, seed = 1)
sampler$step(50)
for (draw in seq_len(100)) {
  sampler$step(1)
  sampler$keep()
}
driven <- sampler$finish()

fit <- thiessen(x, y, control, seed = 1, chains = 1)
identical(predict(driven, x), predict(fit, x))
```

## Imputation between sweeps

A right-censored response can be redrawn from its conditional before
each sweep: read the current mean function and noise level, redraw the
censored values above the limit, and hand the response back. The redraw
below inverts the conditional tail probability, so it is a draw from the
truncated normal. This is the type-I tobit model of Tobin (1958) with
the augmentation of Chib (1992), in the loop rather than as a family.

```{r}
set.seed(1)
latent <- x[, "a"] + rnorm(n, sd = 0.1)
limit <- quantile(latent, 0.8)
censored <- latent > limit
y_observed <- pmin(latent, limit)

sampler <- thiessen_sampler(x, y_observed, control, seed = 1)
sampler$step(50)
for (draw in seq_len(100)) {
  mean_f <- sampler$fitted_values()
  scale <- sqrt(sampler$noise_variances())
  lower <- pnorm(limit, mean_f[censored], scale[censored])
  tail_draw <- qnorm(
    lower + runif(sum(censored)) * (1 - lower),
    mean_f[censored], scale[censored]
  )
  working <- y_observed
  working[censored] <- tail_draw
  sampler$set_response(working)
  sampler$step(1)
  sampler$keep()
}
fit_censored <- sampler$finish()
fit_censored$n_draws
```

The censored fit recovers the upper tail that a fit to the clamped
response flattens at the limit:

```{r}
clamped <- thiessen(x, y_observed, control, seed = 1, chains = 1)
grid <- cbind(a = seq(0, 1, length.out = 100), b = 0.5)
curves <- rbind(
  data.frame(a = grid[, "a"], f = predict(fit_censored, grid),
             fit = "censoring imputed"),
  data.frame(a = grid[, "a"], f = predict(clamped, grid),
             fit = "clamped response")
)
ggplot(curves, aes(a, f, colour = fit)) +
  geom_point(aes(a, y_observed), data.frame(a = x[, "a"], y_observed),
             inherit.aes = FALSE, alpha = 0.3) +
  geom_line() +
  geom_abline(linetype = "dashed") +
  geom_hline(yintercept = limit, linetype = "dotted") +
  labs(y = "f(a, b = 0.5)", colour = NULL)
```

The dashed line is the truth, the dotted one the censoring limit.

## Prototyping an outcome model

This is the proof of the claim at the top of the page: fifteen lines
of R reimplement a shipped outcome family and agree with it. An outcome
model with a latent Gaussian representation can be prototyped in the
loop before it earns a family: impute the latent from its
conditional given the observed outcome, hand it back, sweep. The probit
model is the worked case, its latent drawn from the normal truncated to
the side its label demands (Albert and Chib 1993).

```{r}
set.seed(1)
labels <- as.numeric(x[, "a"] + rnorm(n, sd = 0.3) > 0.5)

z <- ifelse(labels == 1, 0.5, -0.5)
sampler <- thiessen_sampler(x, z, control, seed = 1)
sampler$step(50)
latent_mean <- numeric(n)
for (draw in seq_len(100)) {
  mean_f <- sampler$fitted_values()
  scale <- sqrt(sampler$noise_variances())
  # The conditional mass below zero, then a draw from the label's side
  # of it, by inversion.
  at_zero <- pnorm(0, mean_f, scale)
  u <- runif(n)
  inside <- ifelse(labels == 1, at_zero + u * (1 - at_zero), u * at_zero)
  sampler$set_response(qnorm(inside, mean_f, scale))
  sampler$step(1)
  latent_mean <- latent_mean + sampler$fitted_values()
}
latent_mean <- latent_mean / 100

family <- thiessen(x, labels, thiessen_control(
  outcome = probit_outcome(),
  tessellations = 10,
  general_params = general_params(burn_in = 50, draws = 100)
), seed = 1, chains = 1)
mean((latent_mean > 0) == (predict(family, x) > 0.5))
```

The prototype and the family are the same model up to one difference:
the family fixes the latent variance at 1 for identification, while the
prototype keeps sampling sigma^2. Differences of that kind are what
graduate a prototype into an outcome family. A prototype is checked
distributionally, by simulation-based calibration and posterior
summaries within Monte Carlo error, never bitwise against the family.

## The verbs

`$step(n)` runs n sweeps; `$keep()` records the state as a posterior
draw; `$n_kept()` counts them; `$set_response(y)` replaces the response,
keeping the tessellations, the cell values and sigma^2; `$fitted_values()`
and `$noise_variances()` read the current state at the training rows;
`$finish()` returns the fit `thiessen()` returns and consumes the
sampler. Every later call on a finished sampler errors.

## References

Albert, J. H. and Chib, S. (1993). Bayesian analysis of binary and
polychotomous response data. *Journal of the American Statistical
Association* 88(422), 669-679. doi:10.1080/01621459.1993.10476321

Chib, S. (1992). Bayes inference in the Tobit censored regression
model. *Journal of Econometrics* 51(1-2), 79-99.
doi:10.1016/0304-4076(92)90030-U

Tobin, J. (1958). Estimation of relationships for limited dependent
variables. *Econometrica* 26(1), 24-36. doi:10.2307/1907382
