---
title: "The sampler API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The sampler API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`thiessen_sampler()` is the researcher's interface, and it is
experimental: it may change in a minor release without a deprecation
cycle. It follows the updatable sampler object of dbarts and the
low-level interface of stochtree: construct with the configuration, the
data and a seed, then drive the Gibbs loop yourself. Burn-in and
thinning are the caller's loop, and the response may be replaced between
sweeps. Anything that is not an outcome family or a setting goes through
this loop and stays there until it earns a name.

## The contracts

The response is on the caller's scale through an affine map frozen at
construction, so a response outside the training range is legitimate;
that is what makes censoring and imputation work. The sampler owns its
RNG, seeded at construction with the chain that `thiessen()` would run
first, so driving the configured schedule by hand reproduces a one-chain
fit bit for bit. The loop cannot rewire tessellation membership or cell
internals.

## A fit as its own loop

```{r}
library(thiessen)

set.seed(1)
n <- 120
x <- cbind(runif(n), runif(n))
y <- x[, 1] + rnorm(n, sd = 0.1)
control <- thiessen_control(
  tessellations = 10,
  general_params = general_params(burn_in = 50, draws = 100)
)

sampler <- thiessen_sampler(x, y, control, seed = 1)
sampler$step(50)
for (draw in seq_len(100)) {
  sampler$step(1)
  sampler$keep()
}
driven <- sampler$finish()

fit <- thiessen(x, y, control, seed = 1)
identical(predict(driven, x), predict(fit, x))
```

## Imputation between sweeps

A right-censored response can be redrawn from its conditional before
each sweep: read the current mean function and noise level, redraw the
censored values above the limit, and hand the response back. The redraw
below inverts the conditional tail probability, so it is a draw from the
truncated normal.

```{r}
set.seed(1)
latent <- x[, 1] + rnorm(n, sd = 0.1)
limit <- quantile(latent, 0.8)
censored <- latent > limit
y_observed <- pmin(latent, limit)

sampler <- thiessen_sampler(x, y_observed, control, seed = 1)
sampler$step(50)
for (draw in seq_len(100)) {
  mean_f <- sampler$fitted_values()
  scale <- sqrt(sampler$noise_variances())
  lower <- pnorm(limit, mean_f[censored], scale[censored])
  tail_draw <- qnorm(
    lower + runif(sum(censored)) * (1 - lower),
    mean_f[censored], scale[censored]
  )
  working <- y_observed
  working[censored] <- tail_draw
  sampler$set_response(working)
  sampler$step(1)
  sampler$keep()
}
fit_censored <- sampler$finish()
fit_censored$n_draws
```

The censored fit recovers more of the upper tail than a fit to the
clamped response:

```{r}
clamped <- thiessen(x, y_observed, control, seed = 1)
c(
  censored = max(predict(fit_censored, x)),
  clamped = max(predict(clamped, x)),
  limit = unname(limit)
)
```

## The verbs

`$step(n)` runs n sweeps; `$keep()` records the state as a posterior
draw; `$n_kept()` counts them; `$set_response(y)` replaces the response,
keeping the tessellations, the cell values and sigma^2; `$fitted_values()`
and `$noise_variances()` read the current state at the training rows;
`$finish()` returns the fit `thiessen()` returns and consumes the
sampler. Every later call on a finished sampler errors.
