---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

AddiVortes is Bayesian regression on a sum of Voronoi tessellations
(Stone and Gosling, 2025, *Journal of Computational and Graphical
Statistics* 34(3), 859-871). It stands to BART as a tessellation stands
to a tree: a cell is a region of the covariate space rather than a box,
so a boundary oblique to the axes costs one cell rather than many
splits. This vignette fits the Gaussian model with the defaults; the
other published variants have their own vignettes, by their paper names.

## A first fit

`thiessen()` takes a formula and a data frame, a data frame and a
response, or a numeric matrix and a response. The sweep schedule here is
short so the vignette builds quickly; the defaults are 200 burn-in
sweeps and 1000 kept draws.

```{r}
library(thiessen)

set.seed(1)
n <- 200
d <- data.frame(
  a = runif(n),
  b = runif(n),
  g = factor(sample(c("low", "mid", "high"), n, replace = TRUE))
)
d$y <- 3 * (d$a - 0.4)^2 + 0.5 * d$b + 0.3 * (d$g == "high") +
  rnorm(n, sd = 0.1)

control <- thiessen_control(
  tessellations = 25,
  general_params = general_params(burn_in = 100, draws = 200)
)
fit <- thiessen(y ~ ., d, control, seed = 1)
fit
```

A factor covariate becomes d - 1 treatment-contrast indicators, the
first level as reference, as `model.matrix` encodes one. `predict()`
returns the posterior mean and matches new data by column name and type:

```{r}
head(predict(fit, d))
head(predict(fit, d, interval = "credible", level = 0.9))
```

`predict(interval = "prediction")` covers a new observation rather than
the mean, and `type = "draws"` returns the per-draw values.

`fitted()`, `residuals()`, `sigma()`, `nobs()`, `print()`, `summary()`
and `plot()` behave as they do on any model object, and
`update(fit, seed = 2)` refits with that one argument replaced.

```{r}
c(nobs = nobs(fit), sigma = sigma(fit),
  rmse = sqrt(mean(residuals(fit)^2)))
```

## Covariate scaling

Pass raw data. Inside the sampler a Euclidean column is min-max scaled
to [-0.5, 0.5] over its training range, which is what makes one distance
comparable across columns of different units; a categorical column and a
spherical one are not scaled. The centre-proposal scale `sigma_c` is on
that internal scale, so `sigma_c = 1` is the full training range of a
Euclidean column. Nothing else about a fit depends on the units the
covariates arrive in.

## The configuration

The configuration has four parts, named as the core stores them: an
outcome family from `gaussian_outcome()` or `probit_outcome()`, one
`term_params()` group per ensemble, and the sweep schedule from
`general_params()`.
`thiessen_control(tessellations = )` is the one shortcut, setting the
mean ensemble's size. The control-surface vignette walks through every
group.

```{r}
thiessen_control(
  outcome = gaussian_outcome(nu = 6, q = 0.85),
  mean_params = term_params(tessellations = 200, k = 3, lambda_c = 5)
)
```

## Reproducibility

`seed = NULL`, the default, draws the chain's seed from R's stream, so
`set.seed()` governs; a whole number reproduces the same draws for a
given package version and platform. The resolved seed is on the fit.

```{r}
again <- thiessen(y ~ ., d, control, seed = 1)
identical(predict(fit, d), predict(again, d))
```

## Saving a fit

A fit is a plain R object holding the sampler state, so `saveRDS()`
writes one and a later session reads it and predicts the same values
without a refit.

```{r}
path <- tempfile(fileext = ".rds")
saveRDS(fit, path)
identical(predict(readRDS(path), d), predict(fit, d))
```

## Diagnostics

The established generics are available: `posterior::as_draws_df()`,
`summary()`, and the rstantools generics `posterior_predict()`,
`posterior_epred()`, `log_lik()` and `predictive_interval()`. Two or
more `chains` add rank-normalised split R-hat and effective sample
sizes, and a fit warns where they cross the usual thresholds.

```{r}
summary(fit)
```

`variable_inclusion()` reports the share of the ensemble's active
dimensions falling on each covariate.

```{r}
variable_inclusion(fit)
```

Read that as where the ensemble spent its dimensions, not as variable
selection. At the default `omega` of `min(3, p)` every dimension is
always active when p is 3 or fewer, so the proportions are then uniform
by construction; p is 4 here after the factor encoding, and the
informative covariates separate from the rest only weakly.

`thiessen_diagnostics()` returns the per-draw quantities as a data
frame, for `bayesplot::mcmc_trace()` or a plot of your own. It is not
for printing: a bare `print()` dumps every draw.

## Progress

Progress over the sweep schedule is signalled with progressr, and
nothing is printed by default: the package raises the conditions and the
session decides whether and how to report them. Choose a handler once
and it applies to every fit afterwards.

``` r
progressr::handlers(global = TRUE)
progressr::handlers("txtprogressbar")
fit <- thiessen(y ~ ., d, control, seed = 1)
```

`progressr::with_progress()` scopes a handler to one expression instead.
The chunks here are not evaluated because progressr renders nothing into
a static document. The schedule raises one progression per sweep, to a
maximum of a hundred over the sweeps of every chain, and reporting does
not change the draws.

## Troubleshooting

### The fit prints nothing while it runs

That is the default, not a hang. See Progress above.

### R-hat or an effective sample size warns

`summary()` reports both. More draws is the first answer, then more
chains; `thinning` reduces the stored draws without adding information,
so it is not the remedy for a low effective sample size. The warning
fires on any short schedule, this article's included, and carries the
condition class `thiessen_warning`, so it can be silenced by class
rather than by message:

```{r}
withCallingHandlers(
  two <- thiessen(y ~ ., d, control, chains = 2, seed = 1),
  thiessen_warning = function(condition) {
    invokeRestart("muffleWarning")
  }
)
two$n_chains
```

### A saved fit will not load

A fit written by a build carrying the core's experimental feature and
read by a build without it errors with the class `thiessen_error`,
naming the feature. `core_version()` and `core_experimental()` report
the core version and the feature setting of the build in use; a bug
report should carry both.

```{r}
c(core = core_version(), experimental = core_experimental())
```
